---
title: "Exercise A2"
subtitle: "Statistical Analysis of Big Data"
author: "Amanda Magzal 207608647"
output: pdf_document
header-includes:
  - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
---

```{r include=FALSE}
library(knitr)
```


Consider the following transaction data:

```{r echo=FALSE}
Items <- c("{Bread, Milk}", "{Bread, Diapers, Beer, Eggs}",
           "{Milk, Diapers, Beer, Cola}", "{Bread, Milk, Diapers, Beer}",
           "{Bread, Milk, Diapers, Cola}")
TID <- 1:5
dat <- as.data.frame(cbind(TID, Items))
kable(dat, caption = "Transaction Data", align = "cl")
```

&nbsp;


### (a) Association Rules

&nbsp;

Let $X=\{\text{Milk, Diapers}\}$ and $Y=\{\text{Beer}\}$, with association rule $X \rightarrow Y$.

The support count of itemset $I$ is denoted by $\sigma (I)$.

\underline{Support:}
$$s(X \rightarrow Y) = \frac{\sigma (X \cup Y)}{N}$$

Where $N$ refers to the total number of transactions.

In the given transaction data, 2 out of the 5 transactions contain $X \cup Y = \text{\{Milk, Diapers, Beer\}}$ (ID: 3, 4).

$$s(X \rightarrow Y) = \frac{2}{5} = 0.4$$

\underline{Confidence:}
$$c(X \rightarrow Y) = \frac{\sigma (X \cup Y)}{\sigma (X)}$$

In the given transaction data, the support count for $X$ is 3 (ID: 3, 4, and 5), and he support count for $X \cup Y$ is 2 (ID: 3 and 4). Hence, 2 out of the 3 transactions that contain $X$ also contain $Y$.

$$c(X \rightarrow Y) = \frac{2}{3} = 0.67$$

&nbsp;


### (b) Apriori Algorithm Pseudo-Code

&nbsp;

Let $C_k$ denote the k-itemsets candidates, and $F_k$ denote the frequent k-itemsets.

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\BlankLine
$k = 1$\;
$F_1 = $ find all frequent 1-itemsets\;
\While{$F_k \ne \emptyset$}{
    $k = k+1$\;
    $C_k = $ generate k-itemsets candidates using $F_{k-1}$\;
    $C_k = $ prune candidates using $C_k \text{ and } F_{k-1}$\;
    \For{$\text{each}$ candidate in $C_k$}{
        calculate the support count\;
    }
    $F_k = $ candidates in $C_k$ with support count $> minsup$\;
}
\Return{$\cup _k F_k$}
\caption{Apriori Algorithm}
\end{algorithm}

&nbsp;


### (c) Applying the Apriori Algorithm

&nbsp;

The required $minsup$ is $60\%$, hence the min support count is 3.


1. For each item in the transaction data, calculate the support count and generate the 1-itemsets candidates.

```{r echo=FALSE}
items1 <- c("{Bread}", "{Milk}", "{Diapers}", "{Beer}", "{Eggs}", "{Cola}")
count1 <- c(4, 4, 4, 3, 1, 2)
itemsets1 <- as.data.frame(cbind(items1, count1))
kable(itemsets1, caption = "1-itemsets Candidates", col.names = c("Items", "Support Count"), align = "cc")
```

2. Compare each candidate's support count with the minimum support count. The items $\text{\{Eggs\} and \{Cola\}}$ do not satisfy the minimum support and therefore are not frequent.

```{r echo=FALSE}
items1 <- c("{Bread}", "{Milk}", "{Diapers}", "{Beer}")
count1 <- c(4, 4, 4, 3)
itemsets1 <- as.data.frame(cbind(items1, count1))
kable(itemsets1, caption = "Frequent 1-itemsets", col.names = c("Items", "Support Count"), align = "cc")
```

3. Generate 2-itemsets candidates from the frequent 1-itemsets, and calculate the support count of each item.

```{r echo=FALSE}
items2 <- c("{Bread, Milk}", "{Bread, Diapers}", "{Bread, Beer}", 
            "{Milk, Diapers}", "{Milk, Beer}", "{Diapers, Beer}")
count2 <- c(3, 3, 2, 3, 2, 3)
itemsets2 <- as.data.frame(cbind(items2, count2))
kable(itemsets2, caption = "2-itemsets Candidates", col.names = c("Items", "Support Count"), align = "cc")
```

4. Compare each candidate's support count with the minimum support count.
The items $\text{\{Bread, Beer\} and \{Milk, Beer\}}$ do not satisfy the minimum support and therefore are not frequent.

```{r echo=FALSE}
items2 <- c("{Bread, Milk}", "{Bread, Diapers}", 
            "{Milk, Diapers}", "{Diapers, Beer}")
count2 <- c(3, 3, 3, 3)
itemsets2 <- as.data.frame(cbind(items2, count2))
kable(itemsets2, caption = "Frequent 2-itemsets", col.names = c("Items", "Support Count"), align = "cc")
```

5. Generate 3-itemsets candidates from the frequent 2-itemsets.

```{r echo=FALSE}
items3 <- c("{Bread, Milk, Diapers}", "{Bread, Milk, Beer}",
            "{Bread, Diapers, Beer}", "{Milk, Diapers, Beer}")
itemsets3 <- as.data.frame(items3)
kable(itemsets3, caption = "3-itemsets", col.names = c("Items"), align = "c")
```

Prune the 3-itemsets candidates using the Apriory property - all subsets of frequent items must also be frequent.

- The itemset $\text{\{Bread, Milk, Beer\}}$ includes the subset $\text{\{Milk, Beer\}}$ which is not frequent. Therefore, it cannot be a frequent itemset.

- The itemset $\text{\{Bread, Diapers, Beer\}}$ includes the subset $\text{\{Bread, Beer\}}$ which is not frequent. Therefore, it cannot be a frequent itemset.

- The itemset $\text{\{Milk, Diapers, Beer\}}$ includes the subset $\text{\{Milk, Beer\}}$ which is not frequent. Therefore, it cannot be a frequent itemset.


```{r echo=FALSE}
items3 <- c("{Bread, Milk, Diapers}")
count3 <- c(2)
itemsets3 <- as.data.frame(cbind(items3, count3))
kable(itemsets3, caption = "3-itemsets Candidates", col.names = c("Items", "Support Count"), align = "cc")
```

6. Compare the candidate's support count with the minimum support count. The item $\text{\{Bread, Milk, Diapers\}}$ does not satisfy the minimum support and therefore is not frequent. Hence, there are no frequent 3-itemsets.






